Design and Implementation of a Multithreaded Virtual Machine for Executing Linear Logic Programs

### 2014-08-07 virtual_machine.tex
ricroc's changes.

We have developed a compiler that compiles LM programs to byte-code
and a multi-threaded virtual machine (VM) using POSIX threads
that runs the byte-code.

@@

\begin{figure}[ht]
{\footnotesize
\begin{Verbatim}
void thread_work_loop(thread_id tid)
   while (true)
      node = pop_node_from_work_queue(tid)
      if (node)
         process_node(tid, node)
      else
         // need to steal a node
         target = random(NUM_THREADS)
         for (i = 0; i < NUM_THREADS && !node; i++)
            target = (target + 1) % NUM_THREADS
            node = steal_node_from_thread(target)
         if (!node)
            // try to terminate
            become_idle(tid)
            if (!synchronize_termination(tid))
               return
            become_active(tid)
\end{Verbatim}
}
\caption{Thread work loop}
\label{code:work_loop}
\end{figure}

@@

Figure~\ref{fig:hash_table} shows an example for a hash table data
structure for a \texttt{a(int,int)} predicate with 3 linear facts
indexed by the second argument and stored as a doubly linked list in
bucket \texttt{2}. Each linear fact contains the regular list
pointers, a \texttt{flags} field and the fact arguments. Those are all
stored continuously to improve data locality. One use of
the \texttt{flags} field is to mark that a fact is already being
used. For example, consider the rule \texttt{a(A,B), a(C,D) -o
...}. When we first pick a fact for \texttt{a(A, B)} from the hash
table, we mark it as being used in order to ensure that, when we
retrieve facts for \texttt{a(C, D)}, the first one cannot be used
twice since that would violate linearity.

@@

The rule engine decides which rules may need to be executed while
taking into account rule priorities. Figure~\ref{fig:rule_engine}
shows the rule engine data structures in more detail. There are 4 main
data structures for scheduling rule execution: \texttt{Rule Queue} is
the bitmap representing the rules scheduled to run; \texttt{Active
Bitmap} contains the rules that have enough facts to be
fired; \texttt{Dropped Bitmap} contains the rules that must be dropped
from \texttt{Rule Queue} and \texttt{Active Bitmap} if the rule being
executed succeeds; and \texttt{Predicates Count} counts the number of
facts per predicate. To understand how our engine works, consider the
following set of rules and facts:

{\footnotesize
\begin{Verbatim}
// rules
1: a, e(1) -o b.
2: a -o c.
3: b -o d.
4: e(0) -o f.
5: c -o e(1).

// facts
a.
e(0).
\end{Verbatim}
}

@@

Where $A$ is the target argument, $F$ is the set of linear facts
for the target predicate, $values(F, A)$ is set of values of the
argument $A$, $count(F, A = v)$ counts the number of linear facts
where argument $A$ is equal to $v$ and $total(F)$ counts the number of
linear facts in $F$.  The entropy value is a good metric because it
tells us how much information is needed to describe an argument.  If
more information is needed, then that must be the best argument to
index.

@@

LM also supports recursive types such as lists and pairs. These
complex data structures are stored in the heap of the VM and are
managed through reference counting. For instance, each list is
a \emph{cons cell} with 3 fields: \texttt{tail}, the pointer to the
next element of the list; \texttt{head}, the element stored by this
element of the list; and \texttt{refs} that counts the number of
pointers to this list element in the VM. The list is deleted from the
heap whenever \texttt{refs} is decremented to 0.

>>>>

We have developed a compiler that compiles LM programs to byte-code
and a multithreaded virtual machine (VM) using POSIX threads
that runs the byte-code.

@@

\begin{figure}[ht]
{\footnotesize
\begin{Verbatim}
void thread_work_loop(thread_id tid)
   while (true)
      node = pop_node_from_work_queue(tid)
      if (node)
         process_node(tid, node)
      else
         // need to steal a node
         target = random(NUM_THREADS)
         for (i = 0; i < NUM_THREADS && !node; i++)
            target = (target + 1) % NUM_THREADS
            node = steal_node_from_thread(target)
            if (node) break
         if (!node)
            // try to terminate
            become_idle(tid)
            if (synchronize_termination(tid))
               return
            become_active(tid)
\end{Verbatim}
}
\caption{Thread work loop}
\label{code:work_loop}
\end{figure}

@@

Figure~\ref{fig:hash_table} shows an example for a hash table data
structure for a \texttt{a(int,int)} predicate with 3 linear facts
indexed by the second argument and stored as a doubly linked list in
bucket \texttt{2}. Each linear fact contains the regular list
pointers, a \texttt{flags} field and the fact arguments. Those are all
stored continuously to improve data locality. One use of
the \texttt{flags} field is to mark that a fact is already being
used. For example, consider the rule:

{\footnotesize
\begin{Verbatim}
a(A,B), a(C,D) -o ...
\end{Verbatim}
}

When we first pick a fact for \texttt{a(A, B)} from the hash
table, we mark it as being used in order to ensure that, when we
retrieve facts for \texttt{a(C, D)}, the first one cannot be used
twice since that would violate linearity.

@@

The rule engine decides which rules may need to be executed while
taking into account rule priorities. Figure~\ref{fig:rule_engine}
shows the rule engine data structures in more detail. There are 4 main
data structures for scheduling rule execution: \texttt{Rule Queue} is
the bitmap representing the rules scheduled to run; \texttt{Active
Bitmap} contains the rules that have enough facts to be
fired; \texttt{Dropped Bitmap} contains the rules that must be dropped
from \texttt{Rule Queue} and \texttt{Active Bitmap} if the rule being
executed succeeds; and \texttt{Predicates Count} counts the number of
facts per predicate. To understand how our engine works, consider the
following set of facts and rules:

{\footnotesize
\begin{Verbatim}
a.
e(0).

a, e(1) -o b.  // rule 1
a -o c.        // rule 2
b -o d.        // rule 3
e(0) -o f.     // rule 4
c -o e(1).     // rule 5
\end{Verbatim}
}

@@

\noindent
where $A$ is the target argument, $F$ is the set of linear facts
for the target predicate, $values(F, A)$ is set of values of the
argument $A$, $count(F, A = v)$ counts the number of linear facts
where argument $A$ is equal to $v$ and $total(F)$ counts the number of
linear facts in $F$.  The entropy value is a good metric because it
tells us how much information is needed to describe an argument.  If
more information is needed, then that must be the best argument to
index.

@@

LM also supports recursive types such as lists and pairs. These
complex data structures are stored in the heap of the VM and are
managed through reference counting. For instance, each list is
a \emph{cons cell} with 3 fields: \texttt{tail}, the pointer to the
next element of the list; \texttt{head}, the element stored by this
element of the list; and \texttt{refs} that counts the number of
pointers to this list element in the VM. The list is deleted from the
heap whenever \texttt{refs} is decremented to zero.
